Supplementary Information

### 2020-11-23 doc/manuscript/MEE_manuscript.Rmd
working on EQM and NP feedback, including revamp of a couple paragraphs of intro

```{r eval=FALSE, echo=FALSE}
## Target journal: Methods in Ecology & Evolution (IF 7.09)
## Instructions for authors: https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/author-guidelines
# Research article:  should have a maximum of 6000-7000 words (including tables/figure captions and references) and describe new methods and how they may be used.

@@

**Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change.** 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976] and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016]. 
Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
Yet, dendrochronological methods have been optimized to detect climate signals rather than to predict forest productivity and its climate sensitivity [@klesse_sampling_2018]. 
As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

**To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.** 
Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of ring-width index chronologies to select climate variables--most commonly temperature, precipitation, or a moisture index (PDSI or SPEI)--over the current year and, typically, previous year [e.g., @fritts_tree_1976; @zang_treeclim_2015]. 
While appropriate for identifying the strongest correlation between tree growth and an individual climate variable for purposes such as historical climate reconstructions [e.g., @pederson_is_2013; @maxwell_increased_2017]  and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; but see @foster_predicting_2016]. 
Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@beedlow_importance_2013; @foster_predicting_2016], and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate].
Thus, to model tree growth responses to climate, we need an objective approach to systematically evaluate numerous potential climate variables and time windows in order to select principle climate drivers appropriate for inclusion in multivariate models.

**Traditional dendrochronological methods focus on linear climate responses and the settings in which these occur, potentially missing widespread nonlinearities within forest settings, which have been widely observed at other spatio-temporal scales.**
Over time frames of seconds to days, most biological rates—- from photosynthesis to plant and animal respiration—-display a unimodal relationship to temporal variation in temperature. 
Specifically, these rates generally increase exponentially with temperature up to a point [e.g., @brown_toward_2004], typically reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019], and decrease at higher temperatures [e.g., REFS].
On the other end of the spatio-temporal spectrum, across regional to global climate gradients, annual forest productivity plateaus or decreases at high temperature, $T_{max}$, and potential evapotranspiration, $PET$ [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
Similarly, across global gradients, annual forest productivity increases with precipitation only up to a point, after which it plateaus or decreases [@banbury_morgan_global_nodate]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **??**]. 
Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are rare [@tolwinski-ward_bayesian_2013; @tumajer_increasing_2017], and we therefore know little about what, if any, nonlinearities in the growth responses to interannual variation in climate occur for trees within forest settings.

**Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020], yet traditional methods do not characterize its effect or its potential interactions with climate.** 
For dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for identifying climate signals [(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and smaller trees can be more sensitive to temperature (*Rollinson et al. in review*).
However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity. 
To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ and its potential climate interactions.

>>>>

```{r eval=FALSE, echo=FALSE}
## Target journal: Methods in Ecology & Evolution (IF 7.09)
## Instructions for authors: https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/author-guidelines

## potential reviewers: Jane Foster, M. Evans,  van der Pol,

# Research article:  should have a maximum of 6000-7000 words (including tables/figure captions and references) and describe new methods and how they may be used.

@@

**Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change.** 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976] and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016]. 
Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
Yet, dendrochronological records have been traditionally collected and analyzed to detect climate signals rather than to predict forest productivity and its climate sensitivity [@klesse_sampling_2018]. 
As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

**Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.**
Traditional methods begin by fitting a function (commonly a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and average the residuals across cores to form a species-level ring-width index chronology (*DENDRO_REFS*).
Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous year [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction (*DENDRO_REFS*) or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies fundamentally alters and obfuscates individual-level responses (*REF-NEIL?*). 
Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size (discussed below), canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020].

**Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects that may be widespread within forest settings.** 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
Over time frames of seconds to days, most biological rates—- from photosynthesis to plant and animal respiration—-display a unimodal relationship to temporal variation in temperature. 
Specifically, these rates generally increase exponentially with temperature up to a point [e.g., @brown_toward_2004], typically reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019], and decrease at higher temperatures [e.g., REFS].
On the other end of the spatio-temporal spectrum, across regional to global climate gradients, annual forest productivity plateaus or decreases at high temperature, $T_{max}$, and potential evapotranspiration, $PET$ [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
Similarly, across global gradients, annual forest productivity increases with precipitation only up to a point, after which it plateaus or decreases [@banbury_morgan_global_nodate]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **??**]. 
Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
Furthermore, tree-ring studies do not commonly consider the additive or interactive effects of climate variables on annual growth [but see @meko_seascorr_2011; @foster_predicting_2016]. 
Temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], but their influence may occur over different time windows.

**Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020], yet traditional methods do not characterize its effect or its potential interactions with climate.** 
For dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for identifying climate signals [Fritz 1976?(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and smaller trees can be more sensitive to temperature (*Rollinson et al. in review*).
However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity. 
To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ and its potential climate interactions.
